Variational Inference for Inverse Reinforcement Learning with Gaussian Processes

### 2019-03-24 mpaper/paper.tex
A bit of evaluation

\section{Theoretical Justification} \label{sec:proof}
% TODO: Possible improvements:
% 1. Use limits more often.
% 2. Perturbation lemma can be replaced by continuity of A |-> A^-1.
% 3. epsilon is treated as an arbitrary constant.

@@

\begin{figure*}
  \centering
  \begin{minipage}{.5\textwidth}
    \centering
    \includegraphics[width=0.8\columnwidth]{sage1}
    \includegraphics[width=0.8\columnwidth]{sage2}
  \end{minipage}%
  \begin{minipage}{.5\textwidth}
    \centering
    \includegraphics[width=0.8\columnwidth]{elbo_over_B_short}
  \end{minipage}
  \caption{How a well-defined problem with a correct solutions becomes a
    well-defined problem with an incorrect solution. In each plot, the function
    we are trying to optimise is in blue, and its derivative is in red. The
    plots on the left are for $q(\mathbf{u})$ and its derivatives with respect
    to $B_{1,1}$ (at the top) and $B_{2,1}$ (at the bottom), whereas the plots
    on the right are for $\mathbb{E}[v]$ and its corresponding derivatives.}
  \label{fig:B_gradient}
\end{figure*}

@@

Unfortunately, we are not able to use $\frac{\partial \mathcal{L}}{\partial
  \mathbf{B}}$ in order to optimise $\mathbf{B}$. We illustrate the problem in
Figure \ref{fig:B_gradient}. On the left side of the figure, we plot how
$q(\mathbf{u})$ behaves as a function of a diagonal and a non-diagonal element
of $\mathbf{B}$. Both functions have maximum values that can be attained by
following the corresponding derivatives. However, when these derivatives are
used to estimate $\frac{\partial \mathbb{E}[v]}{\partial \mathbf{B}}$, the
resulting derivatives no longer match their corresponding functions, although
the functions themselves still have optimal values: at or below $1$ for
$B_{1,1}$ and at $0$ for $B_{2,1}$. This leads us to consider two restrictions
of the model in order to investigate its convergence behaviour:
\begin{itemize}
\item In \textbf{Scenario 1}, we remove $\frac{\partial \mathbb{E}[v]}{\partial
    \mathbf{B}}$ from $\frac{\partial \mathcal{L}}{\partial \mathbf{B}}$. This
  essentially optimises $\mathbf{B}$ to match $\Kuu$, because then
  $\frac{\partial \mathcal{L}}{\partial \mathbf{B}}$ becomes
  $-\frac{\partial}{\partial \mathbf{B}} \DKL{q(\mathbf{u})}{p(\mathbf{u})}$,
  i.e., we are optimising $\mathbf{B}$ to minimise the difference between the
  prior and the posterior of $\mathbf{u}$.
\item In \textbf{Scenario 2}, we set $\mathbf{B} = \mathbf{I}_m$, and do not
  optimise it at all.
\end{itemize}

We plot how $\mathcal{L}$ as well as policies $\pi(a_1 \mid s_1)$, $\pi(a_2 \mid
s_3)$, and $\pi(a_1 \mid s_2)$ converge over a number of iterations in Figure
\ref{fig:convergence1}. The first two policies correspond to actions taken in
the set of demonstrations $\mathcal{D}$, so we would expect to see their
probabilities converge to values above $0.5$. Note that due to the stochastic
nature of our MDP model, we do not expect to see any probability reach exactly
$1$. The third policy, however, has no relevant data in $\mathcal{D}$, so the
maximum causal entropy framework would put the probability at around $0.5$.

As we ran the algorithm with two stopping conditions---$300$ iterations and the
$\ell_1$-norm of the change in parameter values being below $0.01$---note that
the algorithm terminated early in Scenario~2 but not in Scenario~1, although
there is clear convergence in both cases. An important difference is that
$\pi(a_1 \mid s_1)$ and $\pi(a_2 \mid s_3)$ converge closer and closer to $1$
with the slightly-restricted model of Scenario~1, but stabilise at $0.6$ in
Scenario~2 due to an important part of the model being completely fixed. In both
cases, $\pi(a_1 \mid s_2)$ converges to $0.5$ as expected.

Figure \ref{fig:convergence2} shows how the parameters of the model converge in
both scenarios. The algorithm seems to converge fine in Scenario~2, but most of
the parameters fail to stabilise in Scenario~1. Most importantly, the diagonal
values of the $\mathbf{B}$ matrix diverge to positive infinity, leading to higher
variance in ELBO estimates seen in Figure \ref{fig:convergence1}. The
non-diagonal values, however, converge just fine. Both $\lambda_0$ and
$\lambda_1$ also diverge to positive infinity, while the elements of $\bm\mu$
diverge, but in `the right' direction: as $\mu_2$ increases while $\mu_1$ and
$\mu_3$ decrease, the policies in Figure \ref{fig:convergence1} converge to
their optimal values.

This leaves us with two models: one that converges to reasonable-but-suboptimal
values, and one that diverges to infinite variance but also provides correct
policies.

\begin{figure*}
  \begin{tikzpicture}[
    /pgfplots/colormap/viridis,
    ellipse2/.style={
      draw,
      ellipse,
      /utils/exec={
        \pgfplotscolormapdefinemappedcolor{#1}},
      fill=mapped color},
    edge2/.style={
      ultra thick,
      /utils/exec={
        \pgfplotscolormapdefinemappedcolor{#1}},
      draw=mapped color}]
    \csvreader[
    no head,
    column count=11
    ]{covariance_and_policy_transformed.csv}{1=\x, 2=\y, 3=\Soneone, 4=\Stwotwo, 5=\Sthreethree, 6=\Stwoone, 7=\Sthreeone, 8=\Sthreetwo, 9=\ponetwo, 10=\ptwoone, 11=\pthreetwo}{%
      \node[ellipse2=\Soneone] (s1) at (3*\x, 2*\y) {$s_1$};
      \node[ellipse2=\Stwotwo] (s2) at (3*\x+1, 2*\y+1) {$s_2$};
      \node[ellipse2=\Sthreethree] (s3) at (3*\x+2, 2*\y) {$s_3$};
      \draw[edge2=\Stwoone] (s1) -- (s2);
      \draw[edge2=\Sthreeone] (s1) to (s3);
      \draw[edge2=\Sthreetwo] (s2) to (s3);
    }
    \draw [<->] (2.5, 7.5) node (yaxis) [above] {$\zeta_2$}
    |- (11.5, 1.5) node (xaxis) [right] {$\zeta_1$};
    \foreach \x in {1, 2, 3} {
      \node (x\x) at (1+3*\x, 1.3) {$\x$};
    }
    \foreach \y in {1, 2, 3} {
      \node (y\y) at (2.3, 0.5+2*\y) {$\y$};
    }
  \end{tikzpicture}
  \begin{tikzpicture}
    \csvreader[
    no head,
    column count=11
    ]{covariance_and_policy_transformed.csv}{1=\x, 2=\y, 3=\Soneone, 4=\Stwotwo, 5=\Sthreethree, 6=\Stwoone, 7=\Sthreeone, 8=\Sthreetwo, 9=\ponetwo, 10=\ptwoone, 11=\pthreetwo}{%
      \begin{axis}[
        ybar=0,
        symbolic x coords={a,b,c},
        ticks=none,
        ymax=1,
        ymin=0,
        x=20,
        bar width=20,
        enlarge x limits=0.25,
        at={(350*\x, 110*\y)},
        height=100,
        bar shift=0pt,
        %legend pos=outer south,
        legend style={
          legend columns=-1,
          at={(1.7,-0.4)}
        }
        ]
        \addplot coordinates {(a, \ponetwo)};
        \addplot coordinates {(b, \ptwoone)};
        \addplot coordinates {(c, \pthreetwo)};
        \draw[dashed] (-50,50) -- (270,50);
        \ifthenelse{\x=2 \AND \y=1}{
          \legend{$\pione$,$\pitwo$,$\pithree$};
        }{}
      \end{axis}
    }
    \draw [<->] (2.3, 8.5) node (yaxis) [above] {$\zeta_2$}
    |- (10, 2) node (xaxis) [right] {$\zeta_1$};
    \foreach \x in {1, 2, 3} {
      \node (x\x) at (1+2.5*\x, 1.7) {$\x$};
    }
    \foreach \y in {1, 2, 3} {
      \node (y\y) at (2.1, 1+2.1*\y) {$\y$};
    }
  \end{tikzpicture}
\end{figure*}

\section{Related Work} % 0.5-1 page

@@

\subsection{Variational Inference}
% none of these approaches really work in our situation because they tend to
% depend on properties of how GPs are used in regression that are no longer true
% in IRL

@@

\begin{itemize}
\item Why my theoretical contributions are important...
\item Reasonable results with convergence: policies do converge to optimal
  values, even at the extremes (i.e., $0$ and $1$).
\item Shown how to eliminate the deterministic training conditional
  assumption---a common weakness of previous approaches.
\item Being able to model rewards using full probability distributions without
  limiting assumptions opens up many new research directions.
\item My excuse for the buggy implementation...
\end{itemize}

>>>>

\section{Theoretical Justification} \label{sec:proof}
% TODO: Possible improvements:
% 1. Use limits more often.
% 2. Avoid saying that epsilon is arbitrary and using phrases like 'sufficiently small'.
% TODO: Q should be lowercase

@@

@@

We plot how $\mathcal{L}$ as well as policies $\pi(a_1 \mid s_1)$, $\pi(a_2 \mid
s_3)$, and $\pi(a_1 \mid s_2)$ converge over a number of iterations in Figure
\ref{fig:convergence1}. The first two policies correspond to actions taken in
the set of demonstrations $\mathcal{D}$, so we would expect to see their
probabilities converge to values near $1$. The third policy, however, has no
data in $\mathcal{D}$ that would guide its value, so the maximum causal entropy
framework would put the probability at around $0.5$.

% TODO: rewrite this
As we ran the algorithm with two stopping conditions---$300$ iterations and the
$\ell_1$-norm of the change in parameter values being below $0.01$---note that
the algorithm terminated early in Scenario~2 but not in Scenario~1, although
there is clear convergence in both cases. An important difference is that
$\pi(a_1 \mid s_1)$ and $\pi(a_2 \mid s_3)$ converge closer and closer to $1$
with the slightly-restricted model of Scenario~1, but stabilise at $0.6$ in
Scenario~2 due to an important part of the model being completely fixed. In both
cases, $\pi(a_1 \mid s_2)$ converges to $0.5$ as expected.

% TODO: rewrite
Figure \ref{fig:convergence2} shows how the parameters of the model converge in
both scenarios. The algorithm seems to converge fine in Scenario~2, but most of
the parameters fail to stabilise in Scenario~1. Most importantly, the diagonal
values of the $\mathbf{B}$ matrix diverge to positive infinity, leading to higher
variance in ELBO estimates seen in Figure \ref{fig:convergence1}. The
non-diagonal values, however, converge just fine. Both $\lambda_0$ and
$\lambda_1$ also diverge to positive infinity, while the elements of $\bm\mu$
diverge, but in `the right' direction: as $\mu_2$ increases while $\mu_1$ and
$\mu_3$ decrease, the policies in Figure \ref{fig:convergence1} converge to
their optimal values.

  % \begin{tikzpicture}[
  %   /pgfplots/colormap/viridis,
  %   ellipse2/.style={
  %     draw,
  %     ellipse,
  %     /utils/exec={
  %       \pgfplotscolormapdefinemappedcolor{#1}},
  %     fill=mapped color},
  %   edge2/.style={
  %     ultra thick,
  %     /utils/exec={
  %       \pgfplotscolormapdefinemappedcolor{#1}},
  %     draw=mapped color}]
  %   \csvreader[
  %   no head,
  %   column count=11
  %   ]{covariance_and_policy_transformed.csv}{1=\x, 2=\y, 3=\Soneone, 4=\Stwotwo, 5=\Sthreethree, 6=\Stwoone, 7=\Sthreeone, 8=\Sthreetwo, 9=\ponetwo, 10=\ptwoone, 11=\pthreetwo}{%
  %     \node[ellipse2=\Soneone] (s1) at (3*\x, 2*\y) {$s_1$};
  %     \node[ellipse2=\Stwotwo] (s2) at (3*\x+1, 2*\y+1) {$s_2$};
  %     \node[ellipse2=\Sthreethree] (s3) at (3*\x+2, 2*\y) {$s_3$};
  %     \draw[edge2=\Stwoone] (s1) -- (s2);
  %     \draw[edge2=\Sthreeone] (s1) to (s3);
  %     \draw[edge2=\Sthreetwo] (s2) to (s3);
  %   }
  %   \draw [<->] (2.5, 7.5) node (yaxis) [above] {$\zeta_2$}
  %   |- (11.5, 1.5) node (xaxis) [right] {$\zeta_1$};
  %   \foreach \x in {1, 2, 3} {
  %     \node (x\x) at (1+3*\x, 1.3) {$\x$};
  %   }
  %   \foreach \y in {1, 2, 3} {
  %     \node (y\y) at (2.3, 0.5+2*\y) {$\y$};
  %   }
  % \end{tikzpicture}

\begin{figure}
  \centering
  \begin{tikzpicture}
    \csvreader[
    no head,
    column count=11
    ]{covariance_and_policy_transformed.csv}{1=\x, 2=\y, 3=\Soneone, 4=\Stwotwo, 5=\Sthreethree, 6=\Stwoone, 7=\Sthreeone, 8=\Sthreetwo, 9=\ponetwo, 10=\ptwoone, 11=\pthreetwo}{%
      \begin{axis}[
        ybar=0,
        symbolic x coords={a,b,c},
        ticks=none,
        ymax=1,
        ymin=0,
        x=20,
        bar width=20,
        enlarge x limits=0.25,
        at={(350*\x, 110*\y)},
        height=100,
        bar shift=0pt,
        legend style={
          legend columns=-1,
          at={(1.7,-0.4)}
        }
        ]
        \addplot coordinates {(a, \ponetwo)};
        \addplot coordinates {(b, \ptwoone)};
        \addplot coordinates {(c, \pthreetwo)};
        \draw[dashed] (-50,50) -- (270,50);
        \ifthenelse{\x=2 \AND \y=1}{
          \legend{$\pione$,$\pitwo$,$\pithree$};
        }{}
      \end{axis}
    }
    \draw [<->] (2.3, 8.5) node (yaxis) [above] {$\zeta_2$}
    |- (10, 2) node (xaxis) [right] {$\zeta_1$};
    \foreach \x in {1, 2, 3} {
      \node (x\x) at (1+2.5*\x, 1.7) {$\x$};
    }
    \foreach \y in {1, 2, 3} {
      \node (y\y) at (2.1, 1+2.1*\y) {$\y$};
    }
  \end{tikzpicture}
  \caption{How policy probabilities change after adding more demonstrations. The
    $x$-axis ($y$-axis) shows how many copies of $\zeta_1$ ($\zeta_2$) were in
    $\mathcal{D}$. Each plot shows the values of three policies after 300
    iterations (averaged out over 10 runs). The bottom (top) of each bar plot
    corresponds to probability 0 (1), and dashed lines mark probability 0.5.}
  \label{fig:policies}
\end{figure}

We would expect stochastic probabilities to converge closer to extreme
values (i.e., 0 and 1) as we add more data to the model. We test this hypothesis
with a small-scale experiment in Figure \ref{fig:policies}. The policy $\pitwo$,
in particular, confirms our prediction. As this policy denotes the probability
of going from $s_2$ to $s_1$, we expect it to stay around $0.5$ as long as we
have equal amounts of demonstrations $\zeta_1$ and $\zeta_2$ (this corresponds
to the plots at coordinates $(1, 1)$, $(2, 2)$, and $(3, 3)$). But as we
increase the number of times state $s_1$ appears in our demonstrations (going
to the right across any row of plots), the model recognises the increasing value
of state $s_1$, and the probability increases. In contrast, going up across the
plots reduces the probability, as state $s_3$ becomes more valuable, and the
model prefers it over state $s_1$. Interestingly, the other two policy
probabilities seem to increase with any added data. This is likely due to the
fact that more demonstrations result in higher gradient values, and thus lead to
better-converged policies.

%The experiments were run with a constant step size of 0.1, discarding iterations where either $\bm\Sigma$ or $\bm\Gamma$ becomes near-singular.

%Diagonal and non-diagonal entries of the covariance matrix are normalised to
%$[0, 1]$ separately, so as to emphasize the variation in each. There were no
%negative entries of the covariance matrix after convergence.

%More data creates bigger gradients---need smaller step size. Dynamic step sizes
%don't work very well because they become zero after a big spike (and spikes do
%happen).
\section{Related Work} % 0.5-1 page

@@

\subsection{Variational Inference}

@@

\begin{itemize}
\item Why my theoretical contributions are important...
\item Reasonable results with convergence: policies do converge to optimal
  values, even at the extremes (i.e., $0$ and $1$).
\item Shown how to eliminate the deterministic training conditional
  assumption---a common weakness of previous approaches.
\item Being able to model rewards using full probability distributions without
  limiting assumptions opens up many new research directions.
\end{itemize}
